---
title: "COVID-19 Posts: A Public Dataset Containing 400+ COVID-19 Blog Posts"
date: "2020-10-28"
authors: 
 - admin
 - Rees Morrison
categories: ["R"]
disable_codefolding: true
links:
 - name: ""
   url: https://github.com/connorrothschild/covid-posts/
   icon_pack: fab
   icon: github
image:
  preview_only: true
tags: ["r", "dataset", "shiny"]
summary: "Making hundreds of COVID-19 blog posts from the R community available for public use."
--- 

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.fullwidth = TRUE,
                      out.width = "100%",
                      dpi = 300)

library(tidyverse)
library(funModeling)
library(ggpmisc) 
library(ggrepel) 
library(readxl) 
library(viridis) 
library(tidytext)
library(reactable)

library(cr)
set_cr_theme()

posts <- readxl::read_excel("./data/postsTopicsRoles1019.xlsx") %>% 
  mutate(CoreRole = ifelse(CoreRole == "NotForProfit", "Nonprofit", CoreRole))
today <- format(Sys.Date(), format="%b. %d")
num_posts <- nrow(posts)
```

Over the last few months, we've been collecting hundreds of COVID-19 blog posts from the R community. Today, we are excited to share this dataset publicly, to help bloggers who want to analyze COVID-19 data by unleashing R and the resources of its community by being able to research such posts. 

So far, we have found and recorded `r num_posts` COVID posts in English. In an effort to encourage others to explore such posts, we've published a [Shiny web app](https://connorrothschild.shinyapps.io/covid-posts/){target="_blank"} which allow users to find the names of the `r posts %>% select(Author) %>% distinct() %>% nrow()` bloggers who wrote those posts, their roles, and their country of focus. The app also lets users interactively search the collection of posts by primary topic, post title, date, and whether the post used a particular mathematical technique or data source. To learn more about the evolution of this dataset, one of the authors (Rees) has published nine articles on Medium. The most recent [Medium article](https://medium.com/@rees_32356/my-previous-nine-articles-have-each-focused-on-a-different-finding-from-an-ever-growing-set-of-e774fb608dbd){target="_blank"} links to the series.

We encourage users to submit their own posts for inclusion, which can be done on this [Google Form](https://docs.google.com/forms/d/e/1FAIpQLScfob3uBoumXIRcSW83T0GhXCfi-KI_psGxIqtl_rQ5VHu9ZQ/viewform?usp=sf_link){target="_blank"}. Our dataset, as well as the code for the Shiny app and this post, is available on [GitHub](https://github.com/connorrothschild/covid-posts){target="_blank"}. If anyone has corrections to the dataset, please write Rees (at) ReesMorrison (dot) com.

The remainder of this post highlights some of the findings from the dataset of COVID-19 posts. As will be made evident by the plots that follow, this is by no means a comprehensive review of *every* COVID-19 R blog post, but rather an overview of the data *that we have found*.

# Analysis

### Posts Over Time

As the pandemic has progressed, fewer bloggers have engaged with COVID-related data, as we notice that blog posts peaked in March of 2020. 

```{r weekly}
month <- seq(as.Date("2020-01-01"),
             as.Date("2020-12-01"),
             by = "1 month")

month_numeric <- lubridate::yday(month) / 365 * 52 + 1
month_label <- lubridate::month(month, label = TRUE)

ggplot(posts, aes(x = Week)) +
  geom_histogram(stat = "count") +
  labs(
    x = element_blank(),
    y = "Number of Blog Posts",
    title = "COVID-19 Blog Posts by the R Community, Over Time",
    subtitle = paste("As of", today),
    caption = "Source: Rees Morrison compilation of COVID-19 blog posts that use R"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(breaks = month_numeric,
                     labels = month_label)
```

Some bloggers have been prolific; many more have been one and done. The plot below shows the names and posts of the 23 bloggers who have so far published at least four posts. For an example of how to read the plot, Tim Churches, at the bottom of the y-axis, has published a total of nine posts, but none after early April.

The color of the points corresponds to the work role of the blogger as explained in the legend at the bottom. It is immediately apparent that professors and academic researchers predominate in this group of bloggers. If you include the postgraduate students, universities writ large account for nearly all of the prolific bloggers.

```{r prolificVer2}
prolific <-
  posts %>% select(LastName, Date) %>% 
  group_by(LastName) %>% 
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>% 
  filter(Count > 4) %>% 
  ungroup()

freqBloggers <- prolific$LastName

graphProlific <-
  posts %>% select(LastName, Date, Role) %>% 
  mutate(Date = lubridate::as_date(Date)) %>% 
  dplyr::filter(LastName %in% freqBloggers)

graphProlific %>% 
  mutate(LastName = fct_reorder(LastName, -desc(Role))) %>%
  ggplot(aes(x = Date, y = LastName, colour = Role)) +
  geom_point(pch = 20, size = 4, alpha = .7) +
  # geom_jitter(pch = 20, size = 4, width = 0.1) +
  labs(
    x = element_blank(),
    y = element_blank(),
    colour = element_blank(),
    title = "The Most Prolific COVID-19 Bloggers",
    subtitle = "Timing of posts for authors with more than four posts",
    caption = "Source: Rees Morrison compilation of COVID-19 blog posts that use R"
  ) +
  scale_x_date() +
  theme(legend.position = "bottom",
        legend.direction = 'horizontal')
```

### Roles of Authors

The bloggers in our dataset describe their work-day roles in a variety of ways. One of the authors (Rees) recorded job roles by standardizing the multitude of terms and descriptions, but it is quite possible that this effort misrepresented what some of these bloggers do for a living. We welcome corrections. 

We've further categorized roles into a broad typology where professions fall into one of five categories: *university*, *corporate*, *professional*, *government*, and *nonprofit*. Those broader categories are represented as columns in the following chart.

```{r authorVer2}
# roles <-
#   read_excel(here::here("data/NamesRolesEmailsTwo.Ver1.xls"), trim_ws = TRUE)
# 
# posts <- left_join(posts, roles[, 1:6])
roleSum <-
  posts %>% group_by(CoreRole, Role) %>% summarise(NumPosts = length(Author)) %>%
  arrange(desc(NumPosts)) %>% ungroup() #30

posts %>% 
  group_by(CoreRole) %>% 
  tally() %>% 
  filter(!is.na(CoreRole)) %>% 
  mutate(CoreRole = fct_reorder(CoreRole, -n)) %>% 
ggplot(aes(x = CoreRole, y = n)) +
  geom_col() +
  labs(
    x = element_blank(),
    y = "Number of Blog Posts",
    title = "Role of Authors that Use R to Analyze COVID-19 Data",
    subtitle = paste("As of", today),
    caption = "Source: Rees Morrison compilation of COVID-19 blog posts that use R"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(legend.position = "bottom",
        legend.direction = 'horizontal')
```

### Data Sources

A greater number of data sources related to COVID-19 will yield richer insights. Combining different datasets can shed new light on an issue, yield improvements, and allow authors to contruct better indices and measures. For that reason, one of the authors (Rees) extracted *dataset information* from our collection of blog posts.

For the most part, bloggers identified the data source they drew on for their analysis. On occasion, we had to manually record the data source, or apply some effort to standardize the 140 data sources.

By far the most prevalent data source is Johns Hopkins University, who early, comprehensively and consistently has set the standard for COVID-19 data collection and dissemination to the public.

```{r datasourcesVer2}
postsConnor <-
 read.csv("./data/postsOct20423allText.csv")

textDF_data <- postsConnor %>%
 mutate(data = str_extract_all(postsConnor$text, pattern = "(?<=\\&Data).*?(?=&&)")) %>%
 unnest(cols = c(data))

source("./scripts/clean_text.R")
```

```{css}
.header {
  text-transform: uppercase;
  font-size: 12px;
  font-weight: 400;
  border-bottom: 2px solid #555;
}
```

```{r dataTable}
textDF_data_gat <-
  textDF_data %>% filter(data %nin% c("Simulation", "No data")) %>%
  group_by(data) %>% summarise(Count = n()) %>%
  arrange(desc(Count)) %>% ungroup() %>% 
  filter(!is.na(data),
         data != "",
         data != " ")

textDF_data_gat %>%
  filter(Count > 3) %>%
  ggplot(aes(x = reorder(data, Count), y = Count)) +
  geom_col() +
  labs(
    x = element_blank(),
    y = "Number of Posts Using Data",
    title = "Most Common Data Sources for COVID-19 Blog Posts",
    subtitle = paste("Excluding sources used less than four times, as of", today),
    caption = "Source: Rees Morrison compilation of COVID-19 blog posts that use R"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  coord_flip()
```

Below, you can look at *all* data sources and how often they were referenced in COVID-19 blog posts.

```{r}
reactable(
  textDF_data_gat,
  compact = TRUE,
  striped = TRUE,
  defaultColDef = colDef(headerClass = "header",
                         na = "–"),
  columns = list(data = colDef(name = 'Data Source',
                               width = 300))
)
```

### Blog Post Topics

It may also be the case that readers want a summary of blogs, or to only look at posts that pertain to a certain topic. Assigning each blog post a primary topic introduces a fair amount of subjectivity, to be sure, but the hope is that these broad topics will help researchers find content and colleagues who share similar interests. 

Here, a balloon plot shows nine categories that the `r num_posts` posts address as their primary topic. Epidemiology leads the way, as might be expected, but quite a few posts seem to use COVID data to showcase something else, or apply R in novel ways.

```{r topicsVer2}
topics <-
  posts %>% dplyr::filter(CoreRole != "NA" &
                            Country != "NA") %>% 
  group_by(CoreRole, Topic) %>% 
  summarise(Count = n()) %>%
  ungroup()

topics$Topic <- ordered(
  topics$Topic,
  levels = topics %>% group_by(Topic) %>% summarise(n = sum(Count)) %>%
    arrange(n) %>% distinct() %>% pull(Topic)
)

topics$CoreRole <- ordered(
  topics$CoreRole,
  levels = topics %>% group_by(CoreRole) %>% summarise(n = sum(Count)) %>%
    arrange(desc(n)) %>% distinct() %>% pull(CoreRole)
)

# add a "row" column which will be the y position in the plot
topics <-
  topics %>% group_by(Topic) %>% mutate(row = cur_group_id()) %>% 
  ungroup() %>% 
# add a "col" column which will be the x position in the plot
 group_by(CoreRole) %>% mutate(col = cur_group_id()) %>% 
  ungroup()

# get character vector of variable names for the x axis. the order is important, hence arrange(col)
vars_x_axis <-
  c(topics %>% arrange(col) %>% select(CoreRole) %>% distinct(CoreRole))$CoreRole

# get character vector of observation names for the y axis.
names_y_axis <-
  c(topics %>% arrange(row) %>% select(Topic) %>% distinct(Topic))$Topic

ggplot(topics,
       aes(
         x = factor(col),
         y = factor(row),
         color = Topic,
         alpha = Count,
         size = Count
       )) +
  geom_label(
    aes(label = Count,
        x = col + 0.3),
    alpha = 1,
    size = 4,
    color = "black",
    label.size = 0,
    family = 'Lato'
  ) +
  geom_point() +
  scale_size_area(max_size = 10) +
  scale_alpha_continuous(range = c(0.5, 1)) +
  # scale_color_manual(values = pal) +
  scale_x_discrete(
    breaks = 1:length(vars_x_axis),
    labels = function(x)
      str_wrap(vars_x_axis, width = 10)
  ) +
  scale_y_discrete(breaks = 1:length(names_y_axis), labels = names_y_axis) +
  labs(title = 'Number of Posts by Topic and Employment Category',
       subtitle = paste('As of', today),
       x = element_blank(),
       y = element_blank()) +
  theme(
    axis.line = element_blank(),
    panel.border = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = 'none',
    text = element_text(colour = 'black',
                        family = 'Lato'),
    axis.text = element_text(size = 10.5),
    title = element_text(size = 14),
    plot.title.position = 'plot',
    axis.ticks = element_blank()
  )   
```

# Concluding Thoughts

We encourage you to use our [Shiny application](https://connorrothschild.shinyapps.io/covid-posts/){target="_blank"} to explore the data for yourself. If you'd like to submit your post to be included, fill out this [Google Form](https://docs.google.com/forms/d/e/1FAIpQLScfob3uBoumXIRcSW83T0GhXCfi-KI_psGxIqtl_rQ5VHu9ZQ/viewform?usp=sf_link){target="_blank"}.

As we note in the footer of the application, the R community is intelligent and produces interesting content, but not all of us are experts when it comes to COVID-19. Engaging with these posts will allow you to better understand the application of R to our current moment, and perhaps provide feedback to post authors. We do not endorse the findings of any particular author and encourage you to find accurate, relevant, and recent information from reputable sources such as the [CDC](https://www.cdc.gov/){target="_blank"} and the [WHO](https://www.who.int/){target="_blank"}. 

As always, thanks for reading.